---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here

demo <- tribble(
  ~REASON,             ~freq,
  "Other",             69,
  "Failure to Comply", 147,
  "Good Progress",     27,
  "Full Course",       1197,
  "Conviction",        177
)


ggplot(data = demo) +
  geom_bar(mapping = aes(x = REASON, y = freq), stat = "identity") +
  labs(title = "Termination Reasons in the East Midlands 2023", x = "Reason", y = "Terminations" ) +
  scale_y_continuous(breaks=seq(0,1250,by=100))


# graph of dismissals without "full course" as it makes it hard to compare all other categories

ggplot(data = filter(demo, REASON != "Full Course")) +
  geom_bar(mapping = aes(x = REASON, y = freq), stat = "identity") +
  labs(title = "Termination Reasons in the East Midlands 2023", x = "Reason", y = "Terminations" ) +
  scale_y_continuous(breaks=seq(0,1250,by=100))

# initial graph of the spread of ages

table(AdjudicationsQ12023$`Age group`)
table(filter(AdjudicationsQ12023, Outcomes == "Proved")$age_group)
age_groups <- tribble(
  ~Reason,             ~freq,
  "15-17",             69,            
  "18-20",             147,
  "21-24",             27,
  "25-29",             1197,
  "30-39",             177,
  "40-49",             3359,
  "50-59",             1372,
  "60-69",             334,
  "70+",               70
)

ggplot(data = age_groups) +
  geom_bar(mapping = aes(x = Reason, y = freq), stat = "identity")
      
  



```


```{r load-data}
# load your data 




```


