---
title: "Prison Demographics"
author: "by Team-Name: Jess Part, Adam Smit, Luke Hardwick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```


## Research Question

Describe your research question here. Include why this question might be of interest.

## I'm trying to link our question with current topics of interest

Prison reform has had huge relevance in recent UK politics. Overcrowding particularly has led to the general belief that the current prison system is unsustainable. Guards have been placed in increasingly unsafe working conditions and as a response prisoners face a huge duration of their time trapped in their cells. Chief Inspector of Prisons, Charlie Taylor, has warned that the prison population crisis “will mean more deprivation, squalor and the risk of further violence”. Inadequate conditions are likely to cause prisoners to be released back into the community without receiving the support that they require to prevent reoffending. Hence contributing further to the issue. We are interested in what causes people to initially end up behind bars. We pose the question of what demographic factors lead to individuals committing crime, and how demographics can influence sentencing. Specifically, we aim to determine how demographic factors are influencing crime in specific counties, how much a certain factor is connected to crime and which factors have the greatest influence in both crime and sentencing. 




## Data

Briefly describe your dataset here - include a description of the key variables you will be investigating. 
Include a citation for your data. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to state the retrieval date.






## Findings

The write-up of your project and findings go here. 

Reasons why demographics aren’t the only variables in crime detection:

Increase in policing, 2019 government manifesto to recruit 20,000 additional police officers has been fulfilled

Traditionally density of police force is spread unevenly between demographics

Although crime may be less prevalent in affluent areas it is also less likely to be detected due to this


Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.



88,225 prisoners in England and Wales 
commonslibrary.parliament.uk
https://commonslibrary.parliament.uk/what-is-the-government-doing-to-reduce-pressure-on-prison-capacity/#:~:text=Current%20prison%20population%20and%20capacity,leading%20to%20increased%20prison%20overcrowding.



