---
title: "Prison Demographics"
author: "by Team-Name: Jess Part, Adam Smit, Luke Hardwick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```


## Research Question


Prison reform has had huge relevance in recent UK politics. Overcrowding particularly has led to the general belief that the current prison system is unsustainable. With a 6% increase in prison population from 2022 to a total of 85,851 prisoners in June 2023. Guards have been placed in increasingly unsafe working conditions and as a response inmates face a huge duration of their time trapped in their cells. Chief Inspector of Prisons, Charlie Taylor, has warned that the prison population crisis “will mean more deprivation, squalor and the risk of further violence”. Inadequate conditions are likely to cause prisoners to be released back into the community without receiving the support that they require to prevent reoffending. Hence contributing further to the issue. We are interested in what causes people to initially end up behind bars. We pose the question of what demographic factors relate to individuals committing crime and how demographics can influence sentencing and the treatment inmates experience in prison.




## Data

Prisoner population statistics (England & Wales), Gov.uk

This data set includes the offender management statistics from January to March 2023 as provided by the UK government. Prison establishments record details for individual inmates using the prison IT system, which is the main source of data for this dataset. This includes information including date of birth, sex, religion, nationality, ethnic origin and offence. Which comprise the demographic factors which we are focusing on in this investigation. The dataset also includes information of sentencing by region within the UK & Wales, which we hope to combine with the mean income of regions to establish connections between poverty, crime and sentencing. In order to obtain information regarding income by region we needed to use an additional dataset. Poverty in the UK: Statistics, which includes percentage of people in relative low income and persistent low income by region.




## Findings

The write-up of your project and findings go here. 

Reasons why demographics aren’t the only variables in crime detection:

.Increase in policing, 2019 government manifesto to recruit 20,000 additional police officers has been fulfilled

.Traditionally density of police force is spread unevenly between demographics, although crime may be less prevalent in affluent areas it is also less likely to be detected due to this.


Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

Ministry of Justice, HM Prison and Probation service, 2023, Offender Management statistics quarterly: January to March 2023, Electronic dataset, Viewed 23/11/2023 <https://www.gov.uk/government/statistics/offender-management-statistics-quarterly-january-to-march-2023>

Office for National Statistics, 2023, Employee Earnings in the UK:2023, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/bulletins/annualsurveyofhoursandearnings/2023>

Office for National Statistics, 2023, Employment in Local Authorities, England and Wales: Census 2021, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/employmentandemployeetypes/articles/employmentinlocalauthoritiesenglandandwales/census2021>

Office for National Statistics, 2021, Population Estimates by ethnic group and religion, England and Wales:2019, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/articles/populationestimatesbyethnicgroupandreligionenglandandwales/2019>

Office for National Statistics, 2022, Divorces in England and Wales: 2021, Electronic dataset, Viewed 28/11/2023 <https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/divorce/bulletins/divorcesinenglandandwales/2021>

Office for National Statistics, 2023, Crime in England and Wales: Police Force Area data Tables, Electronic dataset, Viewed 28/11/2023 <https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/policeforceareadatatables> 

GOV.UK, Department for Environment, Food and Rural affairs, 2023, Population Supplementary Data Tables, Electronic dataset, Viewed 28/11/2023 <https://www.gov.uk/government/statistics/population-statistics-for-rural-england>








