---
title: "Prison Demographics"
author: "by Team-Name: Jess Part, Adam Smit, Luke Hardwick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.




```


## Research Question


Prison reform has had huge relevance in recent UK politics. Overcrowding particularly has led to the general belief that the current prison system is unsustainable. With a 6% increase in prison population from 2022 to a total of 85,851 prisoners in June 2023. Guards have been placed in increasingly unsafe working conditions and as a response inmates face a huge duration of their time trapped in their cells. Chief Inspector of Prisons, Charlie Taylor, has warned that the prison population crisis “will mean more deprivation, squalor and the risk of further violence”. Inadequate conditions are likely to cause prisoners to be released back into the community without receiving the support that they require to prevent reoffending. Hence contributing further to the issue. We are interested in what causes people to initially end up behind bars. We pose the question of what demographic factors relate to individuals committing crime and how demographics can influence sentencing and the treatment inmates experience in prison.




## Data

Prisoner population statistics (England & Wales), Gov.uk

This data set includes the offender management statistics from January to March 2023 as provided by the UK government. Prison establishments record details for individual inmates using the prison IT system, which is the main source of data for this dataset. This includes information including date of birth, sex, religion, nationality, ethnic origin and offence. Which comprise the demographic factors which we are focusing on in this investigation. The dataset also includes information of sentencing by region within the UK & Wales, which we hope to combine with the mean income of regions to establish connections between poverty, crime and sentencing. In order to obtain information regarding income by region we needed to use an additional dataset. Poverty in the UK: Statistics, which includes percentage of people in relative low income and persistent low income by region.




## Findings




## References

Ministry of Justice, HM Prison and Probation service, 2023, Offender Management statistics quarterly: January to March 2023, Electronic dataset, Viewed 23/11/2023 <https://www.gov.uk/government/statistics/offender-management-statistics-quarterly-january-to-march-2023>

Office for National Statistics, 2023, Employee Earnings in the UK:2023, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/earningsandworkinghours/bulletins/annualsurveyofhoursandearnings/2023>

Office for National Statistics, 2023, Employment in Local Authorities, England and Wales: Census 2021, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/employmentandemployeetypes/articles/employmentinlocalauthoritiesenglandandwales/census2021>

Office for National Statistics, 2021, Population Estimates by ethnic group and religion, England and Wales:2019, Electronic dataset, Viewed 28/11/2023
<https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/articles/populationestimatesbyethnicgroupandreligionenglandandwales/2019>

Office for National Statistics, 2022, Divorces in England and Wales: 2021, Electronic dataset, Viewed 28/11/2023 <https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/divorce/bulletins/divorcesinenglandandwales/2021>

Office for National Statistics, 2023, Crime in England and Wales: Police Force Area data Tables, Electronic dataset, Viewed 28/11/2023 <https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/policeforceareadatatables> 
Office for National Statistics, 2019, Recorded crime data at Community Safety Partnership and local authority level, Electronic dataset, Viewed 28/11/2023 <https://www.ons.gov.uk/peoplepopulationandcommunity/crimeandjustice/datasets/recordedcrimedataatcommunitysafetypartnershiplocalauthoritylevel>

GOV.UK, HM Revenue & Customs, 2023, Income and tax by county and region:2020 to 2021, Electronic dataset, Viewed 28/11/2023 
<https://www.gov.uk/government/statistics/income-and-tax-by-county-and-region-2010-to-2011>  

GOV.UK, Department for Environment, Food and Rural affairs, 2023, Population Supplementary Data Tables, Electronic dataset, Viewed 28/11/2023 <https://www.gov.uk/government/statistics/population-statistics-for-rural-england>








